---
title: "Simulations and models of stochastic differential equations"
author: "A.C. Guidoum and K. Boukhetala"
date: "`r Sys.Date()`"
output: 
  knitr:::html_vignette:
    toc: yes
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Simulations and models of stochastic differential equations}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width=6, fig.height=4, fig.path='Figs/', fig.show='hold',
                      warning=FALSE, message=FALSE)
library(Sim.DiffProc)
```


# snssde1d()

Assume that we want to describe the following SDE:

Ito form^[The equivalently of $X_{t}^{\text{mod1}}$ the following Stratonovich SDE: $dX_{t} =  \theta X_{t} \circ dW_{t}$.]:

\begin{equation}\label{eq:05}
dX_{t} = \frac{1}{2}\theta^{2} X_{t} dt + \theta X_{t} dW_{t},\qquad X_{0}=x_{0} > 0
\end{equation}

Stratonovich form:
\begin{equation}\label{eq:06}
dX_{t} =  \frac{1}{2}\theta^{2} X_{t} dt +\theta X_{t} \circ dW_{t},\qquad X_{0}=x_{0} > 0
\end{equation}

In the above $f(t,x)=\frac{1}{2}\theta^{2} x$ and $g(t,x)= \theta x$ ($\theta > 0$), $W_{t}$ is a standard Wiener process. To simulate this models using `snssde1d()` function we need to specify:

- The `drift` and `diffusion` coefficients as R expressions that depend on the state variable `x` and time variable `t`.
- The number of simulation steps `N=1000` (by default: `N=1000`).
- The number of the solution trajectories to be simulated by `M=500` (by default: `M=1`).
- The initial conditions `t0=0`, `x0=10` and end time `T=1` (by default: `t0=0`, `x0=0` and `T=1`).
- The integration step size `Dt=0.001` (by default: `Dt=(T-t0)/N`).
- The choice of process types by the argument `type="ito"` for Ito or `type="str"` for Stratonovich (by default `type="ito"`).
- The numerical method to be used by `method` (by default `method="euler"`).

```{r}
theta = 0.5
f <- expression( (0.5*theta^2*x) )
g <- expression( theta*x )
mod1 <- snssde1d(drift=f,diffusion=g,x0=10,M=500,type="ito") # Using Ito
mod2 <- snssde1d(drift=f,diffusion=g,x0=10,M=500,type="str") # Using Stratonovich 
mod1
mod2
```
Using Monte-Carlo simulations, the following statistical measures (`S3 method`) for class `snssde1d()` can be approximated for the $X_{t}$ process at any time $t$:

* The expected value $\text{E}(X_{t})$ at time $t$, using the command `mean`.
* The variance $\text{Var}(X_{t})$ at time $t$.
* The median $\text{Med}(X_{t})$ at time $t$, using the command `median`.
* The quartile of $X_{t}$ at time $t$, using the command `quantile`.
* The skewness and the kurtosis of $X_{t}$ at time $t$, using the command `skewness` and `kurtosis`.
* The central moments up to order $p$ of $X_{t}$ at time $t$, using the command `moment`.
* The empirical $\alpha \%$ confidence interval of $X_{t}$ at time $t$ (from the $2.5th$ to the $97.5th$ percentile), using the command `bconfint`.


The summary of the results of `mod1` and `mod2` at time $t=1$ of class `snssde1d()` is given by:
```{r}
summary(mod1, at = 1)
summary(mod2, at = 1)
```

Hence we can just make use of the `rsde1d()` function to build our random number generator for the conditional density of the $X_{t}|X_{0}$ ($X_{t}^{\text{mod1}}| X_{0}$ and $X_{t}^{\text{mod2}}|X_{0}$) at time $t = 1$.
```{r}
x1 <- rsde1d(object = mod1, at = 1)  # X(t=1) | X(0)=x0 (Itô SDE)
x2 <- rsde1d(object = mod2, at = 1)  # X(t=1) | X(0)=x0 (Stratonovich SDE)
summary(data.frame(x1,x2))
```
The function `dsde1d()` can be used to show the kernel density estimation for $X_{t}|X_{0}$ at time $t=1$ with log-normal curves:
```{r 01,fig.env='figure*', fig.cap='  '}
mu1 = log(10); sigma1= sqrt(theta^2)  # log mean and log variance for mod1 
mu2 = log(10)-0.5*theta^2 ; sigma2 = sqrt(theta^2) # log mean and log variance for mod2
AppdensI <- dsde1d(mod1, at = 1)
AppdensS <- dsde1d(mod2, at = 1)
plot(AppdensI , dens = function(x) dlnorm(x,meanlog=mu1,sdlog = sigma1))
plot(AppdensS , dens = function(x) dlnorm(x,meanlog=mu2,sdlog = sigma2))
```

In Figure 2, we present the flow of trajectories, the mean path (red lines) of solution of \eqref{eq:05} and \eqref{eq:06}, with their empirical $95\%$ confidence bands, that is to say from the $2.5th$ to the $97.5th$ percentile for each observation at time $t$ (blue lines):

```{r 02,fig.env='figure*', fig.cap=' ',fig.show='hold'}
plot(mod1,plot.type="single",ylab=expression(X^mod1))
lines(time(mod1),mean(mod1),col=2,lwd=2)
lines(time(mod1),bconfint(mod1,level=0.95)[,1],col=4,lwd=2)
lines(time(mod1),bconfint(mod1,level=0.95)[,2],col=4,lwd=2)
legend("topleft",c("mean path",paste("bound of",95,"% confidence")),col=c(2,4),lwd=2,cex=0.8)
plot(mod2,plot.type="single",ylab=expression(X^mod2))
lines(time(mod2),mean(mod2),col=2,lwd=2)
lines(time(mod2),bconfint(mod2,level=0.95)[,1],col=4,lwd=2)
lines(time(mod2),bconfint(mod2,level=0.95)[,2],col=4,lwd=2)
legend("topleft",c("mean path",paste("bound of",95,"% confidence")),col=c(2,4),lwd=2,cex=0.8)
```


[Return to snssde1d()](#snssde1d)

# snssde2d()

The following $2$-dimensional SDE's with a vector of drift and a diagonal matrix of diffusion coefficients:

Ito form:
\begin{equation}\label{eq:09}
\begin{cases}
dX_t = f_{x}(t,X_{t},Y_{t}) dt +  g_{x}(t,X_{t},Y_{t}) dW_{1,t}\\
dY_t = f_{y}(t,X_{t},Y_{t}) dt +  g_{y}(t,X_{t},Y_{t}) dW_{2,t}
\end{cases}
\end{equation}

Stratonovich form:
\begin{equation}\label{eq:10}
\begin{cases}
dX_t = f_{x}(t,X_{t},Y_{t}) dt +  g_{x}(t,X_{t},Y_{t}) \circ dW_{1,t}\\
dY_t = f_{y}(t,X_{t},Y_{t}) dt +  g_{y}(t,X_{t},Y_{t}) \circ dW_{2,t}
\end{cases}
\end{equation}
$W_{1,t}$ and $W_{2,t}$ is a two independent standard Wiener process. To simulate $2d$ models using `snssde2d()` function we need to specify:

- The `drift` (2d) and `diffusion` (2d) coefficients as R expressions that depend on the state variable `x`, `y`  and time variable `t`.
- The number of simulation steps `N` (default: `N=1000`).
- The number of the solution trajectories to be simulated by `M` (default: `M=1`).
- The initial conditions `t0`, `x0` and end time `T` (default: `t0=0`, `x0=c(0,0)` and `T=1`).
- The integration step size `Dt` (default: `Dt=(T-t0)/N`).
- The choice of process types by the argument `type="ito"` for Ito or `type="str"` for Stratonovich (default `type="ito"`).
- The numerical method to be used by `method` (default `method="euler"`).

## Ornstein-Uhlenbeck process and its integral

The Ornstein-Uhlenbeck (OU) process has a long history in physics. Introduced in essence by Langevin in his famous 1908 paper on Brownian motion, the process received a more thorough mathematical examination several decades later by Uhlenbeck and Ornstein (1930). The OU process is understood here to be the univariate continuous Markov process $X_t$. In mathematical terms, the equation is written as an Ito equation:
\begin{equation}\label{eq016}
 dX_t = -\frac{1}{\mu} X_t dt + \sqrt{\sigma} dW_t,\quad X_{0}=x_{0}
\end{equation}
In these equations, $\mu$ and $\sigma$ are positive constants called,
respectively, the relaxation time and the diffusion constant. The time integral of the OU process $X_t$ (or indeed of any process $X_t$) is defined to be the process $Y_t$ that satisfies:
\begin{equation}\label{eq017}
Y_{t} = Y_{0}+\int X_{t} dt \Leftrightarrow dY_t = X_{t} dt ,\quad Y_{0}=y_{0}
\end{equation}
$Y_t$ is not itself a Markov process; however, $X_t$ and $Y_t$ together
comprise a bivariate continuous Markov process. We wish to find the solutions $X_t$ and $Y_t$ to the coupled time-evolution equations:
\begin{equation}\label{eq018}
\begin{cases}
dX_t = -\frac{1}{\mu} X_t dt + \sqrt{\sigma} dW_t\\
dY_t = X_{t} dt
\end{cases}
\end{equation}

We simulate a flow of $500$ trajectories of $(X_{t},Y_{t})$, with integration step size $\Delta t = 0.01$, and using  second Milstein method.

```{r}
x=5;y=0
mu=3;sigma=0.5
fx <- expression(-(x/mu),x)  
gx <- expression(sqrt(sigma),0)
mod2d <- snssde2d(drift=fx,diffusion=gx,Dt=0.01,M=500,x0=c(x,y),method="smilstein")
mod2d
summary(mod2d)
```
For plotting (back in time) using the command `plot`, the results of the simulation are shown in Figure 3.

```{r 03,fig.env='figure*', fig.cap=' '}
plot(mod2d)
```

Take note of the well known result, which can be derived from either this equations. That for any $t > 0$ the OU process $X_t$ and its integral $Y_t$ will be the normal distribution with mean and variance given by:
$$
\begin{cases}
\text{E}(X_{t}) =x_{0} e^{-t/\mu} &\text{and}\quad\text{Var}(X_{t})=\frac{\sigma \mu}{2} \left (1-e^{-2t/\mu}\right )\\
\text{E}(Y_{t}) = y_{0}+x_{0}\mu \left (1-e^{-t/\mu}\right ) &\text{and}\quad\text{Var}(Y_{t})=\sigma\mu^{3}\left (\frac{t}{\mu}-2\left (1-e^{-t/\mu}\right )+\frac{1}{2}\left (1-e^{-2t/\mu}\right )\right )
\end{cases}
$$

Hence we can just make use of the `rsde2d()` function to build our random number for $(X_{t},Y_{t})$ at time $t = 10$.
```{r}
out <- rsde2d(object = mod2d, at = 10) 
summary(out)
```

For each SDE type and for each numerical scheme, the density of $X_t$ and $Y_t$ at time $t=10$ are reported using `dsde2d()` function, see e.g. Figure 4: the marginal density of $X_t$ and $Y_t$ at time $t=10$.


```{r 04,fig.env='figure*', fig.cap='  '}
denM <- dsde2d(mod2d,pdf="M",at =10)
denM
plot(denM, main="Marginal Density")
```

Created using `dsde2d()` plotted in (x, y)-space with `dim = 2`. A `contour` and `image` plot of density obtained from a realization of system $(X_{t},Y_{t})$ at time `t=10`.

```{r 05,fig.env='figure*', fig.cap='  '}
denJ <- dsde2d(mod2d,pdf="J",at =10)
denJ
plot(denJ,display="contour",main="Bivariate Density")
plot(denJ,display="image",drawpoints=TRUE,col.pt="green",cex=0.25,pch=19,main="Bivariate Density")
```

A $3$D plot of the density obtained with: 

```{r 06,fig.env='figure*', fig.cap='  '}
plot(denJ,main="Bivariate Density")
```


[Return to snssde2d()](#snssde2d)

## The stochastic Van-der-Pol equation

The Van der Pol (1922) equation is an ordinary differential equation that can be derived from the Rayleigh differential equation by differentiating and setting $\dot{x}=y$, see Naess and Hegstad (1994); Leung (1995) and for more complex dynamics in Van-der-Pol equation see Jing et al. (2006). It is an equation describing self-sustaining oscillations in which energy is fed into small oscillations and removed from large oscillations. This equation arises in the study of circuits containing vacuum tubes and is given by:
\begin{equation}\label{eq:12}
\ddot{X}-\mu (1-X^{2}) \dot{X} + X = 0
\end{equation}
where $x$ is the position coordinate (which is a function of the time $t$), and $\mu$ is a scalar parameter indicating the nonlinearity and the strength of the damping, to simulate the deterministic equation see Grayling (2014) for more details. Consider stochastic perturbations of the Van-der-Pol equation, and random excitation force of such systems by White noise $\xi_{t}$, with delta-type correlation function $\text{E}(\xi_{t}\xi_{t+h})=2\sigma \delta (h)$
\begin{equation}\label{eq:13}
\ddot{X}-\mu (1-X^{2}) \dot{X} + X = \xi_{t},
\end{equation}
where $\mu > 0$ . It's solution cannot be obtained in terms of elementary functions, even in the phase plane. The White noise $\xi_{t}$ is formally derivative of the Wiener process $W_{t}$. The representation of a system of two first order equations follows the same idea as in the deterministic case by letting $\dot{x}=y$, from physical equation we get the above system:
\begin{equation}\label{eq:14}
\begin{cases}
\dot{X} = Y \\
\dot{Y} = \mu \left(1-X^{2}\right) Y - X + \xi_{t}
\end{cases}
\end{equation}
The system can mathematically explain by a Stratonovitch equations:
\begin{equation}\label{eq:15}
\begin{cases}
dX_{t} = Y_{t} dt \\
dY_{t} = \left(\mu (1-X^{2}_{t}) Y_{t} - X_{t}\right) dt + 2 \sigma \circ dW_{2,t}
\end{cases}
\end{equation}

Implemente in R as follows, with integration step size $\Delta t = 0.01$ and using stochastic Runge-Kutta methods 1-stage.

```{r}
mu = 4; sigma=0.1
fx <- expression( y ,  (mu*( 1-x^2 )* y - x)) 
gx <- expression( 0 ,2*sigma)
mod2d <- snssde2d(drift=fx,diffusion=gx,N=10000,Dt=0.01,type="str",method="rk1")
mod2d
```

```{r 07,fig.env='figure*', fig.cap='  '}
plot2d(mod2d) ## in plane (O,X,Y)
plot(mod2d)   ## back in time
```


[Return to snssde2d()](#snssde2d)

# snssde3d()

The following $3$-dimensional SDE's with a vector of drift and a diagonal matrix of diffusion coefficients:

Ito form:
\begin{equation}\label{eq17}
\begin{cases}
dX_t = f_{x}(t,X_{t},Y_{t},Z_{t}) dt +  g_{x}(t,X_{t},Y_{t},Z_{t}) dW_{1,t}\\
dY_t = f_{y}(t,X_{t},Y_{t},Z_{t}) dt +  g_{y}(t,X_{t},Y_{t},Z_{t}) dW_{2,t}\\
dZ_t = f_{z}(t,X_{t},Y_{t},Z_{t}) dt +  g_{z}(t,X_{t},Y_{t},Z_{t}) dW_{3,t}
\end{cases}
\end{equation}

Stratonovich form:
\begin{equation}\label{eq18}
\begin{cases}
dX_t = f_{x}(t,X_{t},Y_{t},Z_{t}) dt +  g_{x}(t,X_{t},Y_{t},Z_{t}) \circ dW_{1,t}\\
dY_t = f_{y}(t,X_{t},Y_{t},Z_{t}) dt +  g_{y}(t,X_{t},Y_{t},Z_{t}) \circ dW_{2,t}\\
dZ_t = f_{z}(t,X_{t},Y_{t},Z_{t}) dt +  g_{z}(t,X_{t},Y_{t},Z_{t}) \circ dW_{3,t}
\end{cases}
\end{equation}
$W_{1,t}$, $W_{2,t}$ and $W_{3,t}$ is a 3 independent standard Wiener process. To simulate this system using `snssde3d()` function we need to specify:

- The `drift` (3d) and `diffusion` (3d) coefficients as R expressions that depend on the state variables `x`, `y` , `z` and time variable `t`.
- The number of simulation steps `N` (default: `N=1000`).
- The number of the solution trajectories to be simulated by `M` (default: `M=1`).
- The initial conditions `t0`, `x0` and end time `T` (default: `t0=0`, `x0=c(0,0,0)` and `T=1`).
- The integration step size `Dt` (default: `Dt=(T-t0)/N`).
- The choice of process types by the argument `type="ito"` for Ito or `type="str"` for Stratonovich (default `type="ito"`).
- The numerical method to be used by `method` (default `method="euler"`).

## Basic example

Assume that we want to describe the following SDE's (3D) in Ito form:
\begin{equation}\label{eq0166}
\begin{cases}
dX_t = 4 (-1-X_{t}) Y_{t} dt + 0.2 dW_{1,t}\\
dY_t = 4 (1-Y_{t}) X_{t} dt + 0.2 dW_{2,t}\\
dZ_t = 4 (1-Z_{t}) Y_{t} dt + 0.2 dW_{3,t}
\end{cases}
\end{equation}
We simulate a flow of 500 trajectories, with integration step size $\Delta t = 0.001$.

```{r}
fx <- expression(4*(-1-x)*y , 4*(1-y)*x , 4*(1-z)*y) 
gx <- rep(expression(0.2),3)
mod3d <- snssde3d(x0=c(x=2,y=-2,z=-2),drift=fx,diffusion=gx,N=1000,M=500)
mod3d
summary(mod3d)
```
```{r 08,fig.env='figure*', fig.cap='  '}
plot(mod3d,union = TRUE)         ## back in time
plot3D(mod3d,display="persp")    ## in space (O,X,Y,Z)
```

For each SDE type and for each numerical scheme, the marginal density of $X_t$, $Y_t$ and $Z_t$ at time $t=1$ are reported using `dsde3d()` function, see e.g. Figure 8.


```{r 09,fig.env='figure*', fig.cap='  '}
den <- dsde3d(mod3d,pdf="M",at =1)
den
plot(den, main="Marginal Density") 
```
For Joint density for $(X_t,Y_t,Z_t)$ (for more details, see package  [sm](https://cran.r-project.org/package=sm) or  [ks](https://cran.r-project.org/package=ks).)

```{r, eval=FALSE, include=TRUE}
denJ <- dsde3d(mod3d,pdf="J")
denJ
plot(denJ,display="persp")
mtext("Contour surface of kernel density estimate")
```

[Return to snssde3d()](#snssde3d)

## Attractive model for 3D diffusion processes

If we assume that $U_w( x , y , z , t )$, $V_w( x , y , z , t )$ and $S_w( x , y , z , t )$ are neglected and the dispersion coefficient $D( x , y , z )$ is constant. A system becomes (see Boukhetala,1996):
\begin{eqnarray}\label{eq19}
% \nonumber to remove numbering (before each equation)
\begin{cases}
  dX_t = \left(\frac{-K X_{t}}{\sqrt{X^{2}_{t} + Y^{2}_{t} + Z^{2}_{t}}}\right) dt + \sigma dW_{1,t} \nonumber\\
  dY_t = \left(\frac{-K Y_{t}}{\sqrt{X^{2}_{t} + Y^{2}_{t} + Z^{2}_{t}}}\right) dt + \sigma dW_{2,t} \\
  dZ_t = \left(\frac{-K Z_{t}}{\sqrt{X^{2}_{t} + Y^{2}_{t} + Z^{2}_{t}}}\right) dt + \sigma dW_{3,t} \nonumber
\end{cases}
\end{eqnarray}
with initial conditions $(X_{0},Y_{0},Z_{0})=(1,1,1)$, by specifying the drift and diffusion coefficients of three processes $X_{t}$, $Y_{t}$ and $Z_{t}$ as R expressions which depends on the three state variables `(x,y,z)` and time variable `t`, with integration step size `Dt=0.0001`.

```{r}
K = 4; s = 1; sigma = 0.2
fx <- expression( (-K*x/sqrt(x^2+y^2+z^2)) , (-K*y/sqrt(x^2+y^2+z^2)) , (-K*z/sqrt(x^2+y^2+z^2)) ) 
gx <- rep(expression(sigma),3)
mod3d <- snssde3d(drift=fx,diffusion=gx,N=10000,x0=c(x=1,y=1,z=1))
mod3d
```
The results of simulation are shown:

```{r 11,fig.env='figure*', fig.cap='  '}
plot3D(mod3d,display="persp",col="blue")
```

[Return to snssde3d()](#snssde3d)

## Transformation of an SDE one-dimensional

Next is an example of one-dimensional SDE driven by three independent Brownian motions ($W_{1,t}$,$W_{2,t}$,$W_{3,t}$), as follows:
\begin{equation}\label{eq20}
  dX_{t} = \mu W_{1,t} dt + \sigma W_{2,t} dW_{3,t}
\end{equation}
To simulate the solution of the process $X_t$, we make a transformation to a system of three equations as follows:
\begin{eqnarray}\label{eq21}
\begin{cases}
% \nonumber to remove numbering (before each equation)
  dX_t =  \mu Y_{t}  dt + \sigma Z_{t} dW_{3,t} \nonumber\\
  dY_t =  dW_{1,t} \\
  dZ_t =  dW_{2,t} \nonumber
\end{cases}
\end{eqnarray}
run by calling the function `snssde3d()` to produce a simulation of the solution, with $\mu = 1$ and $\sigma = 1$.

```{r}
fx <- expression(y,0,0) 
gx <- expression(z,1,1)
modtra <- snssde3d(drift=fx,diffusion=gx,M=500)
modtra
summary(modtra)
```
the following code produces the result in Figure 9.
```{r 12,fig.env='figure*', fig.cap='  '}
plot(modtra$X,plot.type="single",ylab="X")
lines(time(modtra),mean(modtra)$X,col=2,lwd=2)
lines(time(modtra),bconfint(modtra,level=0.95)$X[,1],col=4,lwd=2)
lines(time(modtra),bconfint(modtra,level=0.95)$X[,2],col=4,lwd=2)
legend("topleft",c("mean path",paste("bound of",95,"% confidence")),col=c(2,4),lwd=2,cex=0.8)
```

The histogram and kernel density of $X_t$ at time $t=1$ are reported using `dsde3d()` function, see e.g. Figure 10.

```{r 13,fig.env='figure*', fig.cap='  '}
den <- dsde3d(modtra,at=1)
den$resx
MASS::truehist(den$ech$x,xlab = expression(X[t==1]));box()
lines(den$resx,col="red",lwd=2)
legend("topleft",c("Distribution histogram","Kernel Density"),inset =.01,pch=c(15,NA),lty=c(NA,1),col=c("cyan","red"),lwd=2,cex=0.8)

```


[Return to snssde3d()](#snssde3d)
