---
title: "The Sim.DiffProc Package"
author: 
- A.C. Guidoum^[Department of Probabilities & Statistics, Faculty of Mathematics, University of Science and Technology Houari Boumediene, BP 32 El-Alia, U.S.T.H.B, Algeria, E-mail  (acguidoum@usthb.dz)] and K. Boukhetala^[Faculty of Mathematics, University of Science and Technology Houari Boumediene, BP 32 El-Alia, U.S.T.H.B, Algeria, E-mail (kboukhetala@usthb.dz)]
date: "`r Sys.Date()`"
output: 
  knitr:::html_vignette:
    toc: yes
fontsize: 12pt
vignette: >
  %\VignetteIndexEntry{The Sim.DiffProc Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{latin1}
---

```{r setup, echo = F, message = F, results = 'hide'}
library(Sim.DiffProc)
library(knitr)
knitr::opts_chunk$set(comment="",prompt=TRUE, fig.show='hold',warning=FALSE, message=FALSE)
options(prompt="R> ",scipen=16,digits=5,warning=FALSE, message=FALSE,width = 70)
```

# Package Overview

The package [Sim.DiffProc](https://cran.r-project.org/package=Sim.DiffProc) is an object created in R for symbolic and numerical computations on scalar and multivariate systems of stochastic differential equations. It provides users with a wide range of tools to simulate, estimate, analyze, and visualize the dynamics of these systems in both forms Itô and Stratonovich. The project was officially launched in September 2010 and is under active development by the authors. The current feature set of the package can be split in more main categories: Computing the stochastic integrals of Itô or Stratonovich type. Simulation sde's and bridge sde's of Itô or Stratonovich type (1,2 and 3-dim), with different methods. Approximate transition density and random number generators for SDE's. Density approximation for First-passage-time (f.p.t) in SDE's (1,2 and 3-dim). Statistical analysis with Parallel Monte-Carlo and moment equations methods of SDE's (1,2 and 3-dim). Estimate drift and diffusion parameters using pseudo-maximum likelihood estimators of 1-dim SDE's. Displaying an object inheriting from a class of SDE's.

The package includes the following categories (where `k=1,2,3`):

1. [`snssdekd()` & `dsdekd()` & `rsdekd()`- Monte-Carlo Simulation and Analysis of Stochastic Differential Equations](snssde.html).
2. [`bridgesdekd()` & `dsdekd()` & `rsdekd()` - Constructs and Analysis of Bridges Stochastic Differential Equations](bridgesde.html).
3. [`fptsdekd()` & `dfptsdekd()` - Monte-Carlo Simulation and Kernel Density Estimation of First passage time](fptsde.html).
4. [`MCM.sde()` & `MEM.sde()` - Parallel Monte-Carlo and Moment Equations for SDEs](mcmsde.html).
5. [`fitsde()` - Parametric Estimation of 1-D Stochastic Differential Equation](fitsde.html).


# Documentation

It is a requirement of the R packaging system that every function and data set 
in a package has a help page. The [Sim.DiffProc](https://cran.r-project.org/package=Sim.DiffProc) package follows this 
requirement strictly. In addition to the help pages, the package includes vignettes and demonstration scripts; 

```{r, eval=FALSE, include=TRUE}
browseVignettes(package = "Sim.DiffProc")
```
and 
```{r, eval=FALSE, include=TRUE}
demo(package = "Sim.DiffProc")
```

# Collaboration and citation

Obviously, the package leaves many other fields of stochastic modeling with Itô and Stratonovich SDE's untouched. For this situation to change, we hope that experts in their field will join their efforts to ours and contribute code to the Sim.DiffProc project. The project will continue to grow and improve by the authors to the community of developers and users. If you use [Sim.DiffProc](https://cran.r-project.org/package=Sim.DiffProc) please cite the software in publications;
use `citation()` for information on how to cite the software;
```{r }
citation("Sim.DiffProc")
```

# Note

Please send comments, error reports, etc. to the author via the addresses email.

# References

1. Guidoum AC, Boukhetala K (2017). Performing Parallel Monte Carlo and Moment Equations Methods for Itô and Stratonovich Stochastic Differential Systems: R Package Sim.DiffProc. Preprint submitted to Journal of Statistical Software.

2. Guidoum AC, Boukhetala K (2017). Sim.DiffProc: Simulation of Diffusion Processes. R package version 3.8, URL https://cran.r-project.org/package=Sim.DiffProc.

