---
title: "First-passage-time for stochastic differential equations"
author: "A.C. Guidoum and K. Boukhetala"
date: "`r Sys.Date()`"
output: 
  knitr:::html_vignette:
    toc: yes
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{First-passage-time for stochastic differential equations}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width=6, fig.height=4, fig.path='Figs/', fig.show='hold',
                      warning=FALSE, message=FALSE)
library(Sim.DiffProc)
```


# The fptsde function

A new algorithm based on the Monte Carlo technique to generate the random variable FPT of a time homogeneous diffusion process (1, 2 and 3D) through a time-dependent boundary, order to estimate her probability density function.


Let \(X_t\) be a diffusion process which is the unique solution of the
following stochastic differential equation:

\begin{equation}\label{eds01}
  dX_t = \mu(t,X_t) dt + \sigma(t,X_t) dW_t,\quad X_{t_{0}}=x_{0}
\end{equation}

if \(S(t)\) is a time-dependent boundary, we are interested in
generating the first passage time (FPT) of the diffusion process through
this boundary that is we will study the following random variable:

\[
\tau_{S(t)}=
\left\{
  \begin{array}{ll}
    inf \left\{t: X_{t} \geq S(t)|X_{t_{0}}=x_{0} \right\} & \hbox{if} \quad x_{0} \leq S(t_{0}) \\
    inf \left\{t: X_{t} \leq S(t)|X_{t_{0}}=x_{0} \right\} & \hbox{if} \quad x_{0} \geq S(t_{0})
  \end{array}
\right.
\]

The main arguments to 'random' `rfptsdekd()` (where `k=1,2,3`) consist: 

- `object` an object inheriting from class `snssde1d`, `snssde2d` and `snssde3d`.
- `boundary` an expression of a constant or time-dependent boundary $S(t)$.

The main arguments to 'density' `dfptsdekd()` (where `k=1,2,3`) consist: 

- `object` an object inheriting from class `snssde1d`, `snssde2d` and `snssde3d`.
- `boundary` an expression of a constant or time-dependent boundary $S(t)$.
- `pdf` probability density function `Joint` or `Marginal`.

# Examples

## FPT for 1-Dim SDE

Consider the following SDE and linear boundary:

\begin{align*}
  dX_{t}= & \frac{1}{2} \alpha^2 X_{t} dt + \alpha X_{t} dW_{t},~x_{0} \neq 0.\\
  S(t)= & a+bt
\end{align*}

The analytical solution of this model is: \[
X_t = x_{0}\exp\left(\alpha W_{t}\right)
\] generating the first passage time (FPT) of this model through this
boundary: \[
\tau_{S(t)}=
    \inf \left\{t: X_{t} \geq S(t) |X_{t_{0}}=x_{0} \right\} ~~ \text{if} \quad x_{0} \leq S(t_{0})
\]

Set the model $X_t$:
```{r}
alpha=2
f <- expression( alpha^2 * x )
g <- expression( alpha * x )
mod1d <- snssde1d(drift=f,diffusion=g,x0=0.5,M=1000)
```
Generate the first-passage-time $\tau_{S(t)}$, with `rfptsde1d()` function:
```{r}
St  <- expression( -5*t+1 )
fpt1d <- rfptsde1d(mod1d, boundary = St)
summary(fpt1d)
```

The kernel density of $\tau_{S(t)}$ with boundary $S(t) = -5t+1$, using `dfptsde1d()` function,  see e.g. Figure 2. 

```{r 1,fig.env='figure*', fig.cap='  '}
den <- dfptsde1d(mod1d, boundary = St, bw ='ucv')
den 
plot(den)
```

## FPT for 2-Dim SDE's

The following $2$-dimensional SDE's with a vector of drift and a diagonal matrix of diffusion coefficients:

\begin{equation}\label{eq:09}
\begin{cases}
dX_t = f_{x}(t,X_{t},Y_{t}) dt +  g_{x}(t,X_{t},Y_{t}) dW_{1,t}\\
dY_t = f_{y}(t,X_{t},Y_{t}) dt +  g_{y}(t,X_{t},Y_{t}) dW_{2,t}
\end{cases}
\end{equation}

$W_{1,t}$ and $W_{2,t}$ is a two independent standard Wiener process. First passage time (2D) \((\tau_{(S(t),X_{t})},\tau_{(S(t),Y_{t})})\) is
defined as:

\[
\left\{
  \begin{array}{ll}
    \tau_{S(t),X_{t}}=\inf \left\{t: X_{t} \geq S(t)|X_{t_{0}}=x_{0} \right\} & \hbox{if} \quad x_{0} \leq S(t_{0}) \\
    \tau_{S(t),Y_{t}}=\inf \left\{t: Y_{t} \geq S(t)|Y_{t_{0}}=y_{0} \right\} & \hbox{if} \quad y_{0} \leq S(t_{0})
  \end{array}
\right.
\] and \[
\left\{
  \begin{array}{ll}
   \tau_{S(t),X_{t}}= \inf \left\{t: X_{t} \leq S(t)|X_{t_{0}}=x_{0} \right\} & \hbox{if} \quad x_{0} \geq S(t_{0}) \\
   \tau_{S(t),Y_{t}}= \inf \left\{t: Y_{t} \leq S(t)|Y_{t_{0}}=y_{0} \right\} & \hbox{if} \quad y_{0} \geq S(t_{0})
  \end{array}
\right.
\]

Assume that we want to describe the following SDE's (2D):

\begin{equation}\label{eq016}
\begin{cases}
dX_t = 5 (-1-Y_{t}) X_{t} dt + 0.5 dW_{1,t}\\
dY_t = 5 (-1-X_{t}) Y_{t} dt + 0.5 dW_{2,t}
\end{cases}
\end{equation}

and \[
S(t)=-3+5t
\]

Set the system $(X_t , Y_t)$:
```{r}
fx <- expression(5*(-1-y)*x , 5*(-1-x)*y)
gx <- rep(expression(0.5),2)
mod2d <- snssde2d(drift=fx,diffusion=gx,x0=c(x=2,y=-2),M=1000)
```
Generate the couple \((\tau_{(S(t),X_{t})},\tau_{(S(t),Y_{t})})\), with `rfptsde2d()` function::
```{r}
St <- expression(-3+5*t)
fpt2d <- rfptsde2d(mod2d, boundary = St)
summary(fpt2d)
```

The marginal density of \((\tau_{(S(t),X_{t})},\tau_{(S(t),Y_{t})})\) are reported using `dfptsde2d()` function, see e.g. Figure 4.

```{r 2,fig.env='figure*', fig.cap='  '}
denM <- dfptsde2d(mod2d, boundary = St, pdf = 'M')
denM
plot(denM)
```
A `contour` and `image` plot of density obtained from a realization of system \((\tau_{(S(t),X_{t})},\tau_{(S(t),Y_{t})})\).

```{r 3,fig.env='figure*', fig.cap='  '}
denJ <- dfptsde2d(mod2d, boundary = St, pdf = 'J')
denJ
plot(denJ,display="contour",main="Bivariate Density")
plot(denJ,display="image",drawpoints=TRUE,col.pt="green",cex=0.25,pch=19,main="Bivariate Density")
```

A $3$D plot of the Joint density with: 

```{r 4, echo=TRUE, fig.cap='  ', fig.env='figure*', message=FALSE, warning=FALSE}
plot(denJ,display="persp",main="Bivariate Density")
```

## FPT for 3-Dim SDE's

The following $3$-dimensional SDE's with a vector of drift and a diagonal matrix of diffusion coefficients:

Ito form:
\begin{equation}\label{eq17}
\begin{cases}
dX_t = f_{x}(t,X_{t},Y_{t},Z_{t}) dt +  g_{x}(t,X_{t},Y_{t},Z_{t}) dW_{1,t}\\
dY_t = f_{y}(t,X_{t},Y_{t},Z_{t}) dt +  g_{y}(t,X_{t},Y_{t},Z_{t}) dW_{2,t}\\
dZ_t = f_{z}(t,X_{t},Y_{t},Z_{t}) dt +  g_{z}(t,X_{t},Y_{t},Z_{t}) dW_{3,t}
\end{cases}
\end{equation}
$W_{1,t}$, $W_{2,t}$ and $W_{3,t}$ is a 3 independent standard Wiener process. First passage time (3D) $(\tau_{(S(t),X_{t})},\tau_{(S(t),Y_{t})},\tau_{(S(t),Z_{t})})$ is defined as:

$$
\left\{
  \begin{array}{ll}
    \tau_{S(t),X_{t}}=\inf \left\{t: X_{t} \geq S(t)|X_{t_{0}}=x_{0} \right\} & \hbox{if} \quad x_{0} \leq S(t_{0}) \\
    \tau_{S(t),Y_{t}}=\inf \left\{t: Y_{t} \geq S(t)|Y_{t_{0}}=y_{0} \right\} & \hbox{if} \quad y_{0} \leq S(t_{0}) \\
 \tau_{S(t),Z_{t}}=\inf \left\{t: Z_{t} \geq S(t)|Z_{t_{0}}=z_{0} \right\} & \hbox{if} \quad z_{0} \leq S(t_{0})
  \end{array}
\right.
$$
and
$$
\left\{
  \begin{array}{ll}
   \tau_{S(t),X_{t}}= \inf \left\{t: X_{t} \leq S(t)|X_{t_{0}}=x_{0} \right\} & \hbox{if} \quad x_{0} \geq S(t_{0}) \\
   \tau_{S(t),Y_{t}}= \inf \left\{t: Y_{t} \leq S(t)|Y_{t_{0}}=y_{0} \right\} & \hbox{if} \quad y_{0} \geq S(t_{0}) \\
   \tau_{S(t),Z_{t}}= \inf \left\{t: Z_{t} \leq S(t)|Z_{t_{0}}=z_{0} \right\} & \hbox{if} \quad z_{0} \geq S(t_{0}) \\
  \end{array}
\right.
$$

Assume that we want to describe the following SDE's (3D):
\begin{equation}\label{eq0166}
\begin{cases}
dX_t = 4 (-1-X_{t}) Y_{t} dt + 0.2 dW_{1,t}\\
dY_t = 4 (1-Y_{t}) X_{t} dt + 0.2 dW_{2,t}\\
dZ_t = 4 (1-Z_{t}) Y_{t} dt + 0.2 dW_{3,t}
\end{cases}
\end{equation}
and 
$$
S(t)=-3+5t
$$

Set the system $(X_t , Y_t , Z_t)$:
```{r}
fx <- expression(4*(-1-x)*y , 4*(1-y)*x , 4*(1-z)*y) 
gx <- rep(expression(0.2),3)
mod3d <- snssde3d(drift=fx,diffusion=gx,x0=c(x=2,y=-2,z=0),M=1000)
```
Generate the triplet $(\tau_{(S(t),X_{t})},\tau_{(S(t),Y_{t})},\tau_{(S(t),Z_{t})})$, with `rfptsde3d()` function::
```{r}
St <- expression(-3+5*t)
fpt3d <- rfptsde3d(mod3d, boundary = St)
summary(fpt3d)
```
The marginal density of $(\tau_{(S(t),X_{t})},\tau_{(S(t),Y_{t})},\tau_{(S(t),Z_{t})})$ are reported using `dfptsde3d()` function, see e.g. Figure 4.

```{r 5,fig.env='figure*', fig.cap='  '}
denM <- dfptsde3d(mod3d, boundary = St)
denM
plot(denM)
```

For Joint density for $(\tau_{(S(t),X_{t})},\tau_{(S(t),Y_{t})},\tau_{(S(t),Z_{t})})$ see package [sm](https://cran.r-project.org/package=sm) or  [ks](https://cran.r-project.org/package=ks).

```{r, eval=FALSE, include=TRUE}
library(sm)
sm.density(fpt3d,display="rgl")

##

library(ks)
fhat <- kde(x=fpt3d)
plot(fhat, drawpoints=TRUE)
```
